---
title: "Hands on Exercise 1"
author: "Guan Jhen Lin"
date: "Apr 13, 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Getting Started

<!-- Google tag (gtag.js) -->

\<script async src="https://www.googletagmanager.com/gtag/js?id=G-65B5RWMGTX"\>\</script\> \<script\> window.dataLayer = window.dataLayer \|\| \[\]; function gtag(){dataLayer.push(arguments);} gtag('js', new Date());

gtag('config', 'G-65B5RWMGTX'); \</script\>

Load and Install R packages

```{r}
pacman::p_load(tidyverse, forcats)
```

# Import Data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
summary(exam_data)
```

#Introductions to ggplot The aesthetic mappings take attributes of the data and and use them to influence visual characteristics, such as position, colour, size, shape, or transparency. Each visual characteristic can thus encode an aspect of the data and be used to convey information.

Bar chart

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```

Dot Plot

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,         
               dotsize = 0.5) +      
  scale_y_continuous(NULL,           
                     breaks = NULL)  
```

Histogram

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram()       
```

changing aes()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           fill = GENDER)) +
  geom_histogram(bins=20, 
                 color="grey30") 
```

# ggplot2: geom

geom-density()

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_density()         
```

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS, 
           colour = GENDER)) +
  geom_density()        
```

geom_boxplot

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS,       
           x= GENDER)) +    
  geom_boxplot()  
```

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot(notch=TRUE)
```

geom_violin

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_violin()
```

geom_point()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point()     
```

combining geom objects

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot() +                    
  geom_point(position="jitter", 
             size = 0.5)        
```

# Statistics Functions

The [Statistics](https://ggplot2.tidyverse.org/reference/#stats) functions statistically transform data, usually as some form of summary.

Using stat_summary() function to over ride the default geom.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  geom_point(stat="summary",        
             fun.y="mean",           
             colour ="red",          
             size=4)     
```

Adding best curve on a scatterplot using geom_smooth

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(size=0.5)
```

Without using geom_smooth

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point()
```

Overriding default smoothing method

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5)
```

# ggplot2: facets

facet_wrap()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)
```

facet_grid()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS)
```

# ggplot 2: coordinates

By the default, the bar chart of ggplot2 is in vertical form. The code chunk below flips the horizontal bar chart into vertical bar chart by using coord_flip().

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()
```

Changing the y- and x- axis range

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5)
```

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
```

# Using Themes

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()
```

**Lesson Slides and Hands-on Notes**

-   [Lesson 1: Introduction to Visual Analytics](https://isss608-ay2023-24apr.netlify.app/lesson/lesson01/lesson01-vaa).

-   [Hands-on Exercise 1: A Layered Grammar of Graphics: ggplot2 methods](https://r4va.netlify.app/chap01).

**Self-reading Before Lesson**

**Webinars**

-   [The best Hans Rosling talks you’ve ever seen](https://www.ted.com/talks/hans_rosling_the_best_stats_you_ve_ever_seen?referrer=playlist-the_best_hans_rosling_talks_yo).

-   [VIVA – What is Visual Analytics?](https://www.youtube.com/watch?v=LYKO_6KPZCM/)

-   [Why Visual Analytics](https://www.youtube.com/watch?v=5uGRGqCFryg/)

-   [Visual Analytics - Mastering the Information Age](https://www.youtube.com/watch?v=5i3xbitEVfs/)

**Readings**

-   Thomas, J.J. & Cook, K.A. (2006) “[A Visual Analytics Agenda](https://www-computer-org.libproxy.smu.edu.sg/csdl/magazine/cg/2006/01/mcg2006010010/13rRUxBa5ej)”, ‘’IEEE Computer Graphics and Applications’’, January 2006, Vol.26(1), pp.10-13

-   Chabot, C (2009) “[Demystifying Visual Analytics](https://www-computer-org.libproxy.smu.edu.sg/csdl/magazine/cg/2009/02/mcg2009020084/13rRUy3gn3z)”, ‘’IEEE Computer Graphics and Applications’’, March 2009, Vol.29(2), pp.84-87.

-   [The 5 Most Influential Data Visualizations of All Time](https://www.tableau.com/sites/default/files/whitepapers/the_5_most_influential_data_visualizations_of_all_time.pdf)

-   J.J. Thomas and K.A. Cook, eds. (2005) [Illuminating the Path: The Research and Development Agenda for Visual Analytics](https://ils.unc.edu/courses/2017_fall/inls641_001/books/RD_Agenda_VisualAnalytics.pdf), IEEE CS Press.

-   Jun, Yuan et. al. (2020) “A survey of visual analytics techniques for machine learning”, *Computational Visual Media*. [doi.org/10.1007/s41095-020-0191-7](https://doi.org/10.1007/s41095-020-0191-7).

-   Cui, Wenqiang (2019) “[Visual Analytics: A Comprehensive Overview](https://ntnuopen.ntnu.no/ntnu-xmlui/bitstream/handle/11250/2608888/08740868.pdf?sequence=2)”, IEEE Access.

**Statistical Graphics Methods**

**Age-sex Pyramid**

-   [Population pyramid](https://en.wikipedia.org/wiki/Population_pyramid)

-   [Age-Sex and Population Pyramids: The Most Useful Graphs in Population Geography](https://www.thoughtco.com/age-sex-pyramids-and-population-pyramids-1435272)

-   [Population Pyramid Singapore](https://www.populationpyramid.net/singapore/2017/)

**Pareto Chart**

-   [Pareto chart](https://en.wikipedia.org/wiki/Pareto_chart)

-   [Revising the Pareto Chart](https://www-tandfonline-com.libproxy.smu.edu.sg/doi/abs/10.1198/000313006X152243)

**Trellis Display**

-   [The Visual Design and Control of Trellis Display](https://www-jstor-org.libproxy.smu.edu.sg/stable/1390777?seq=1#metadata_info_tab_contents)

-   [Trellis display](https://onlinelibrary-wiley-com.libproxy.smu.edu.sg/doi/full/10.1002/wics.121)

-   [Trellis Displays vs. Interactive Graphics](https://www.researchgate.net/publication/2671674_Trellis_Displays_vs_Interactive_Graphics)

**All About Tableau**

**Things about Tableau you must know!**

-   [Polaris: A System for Query, Analysis,and Visualization of Multidimensional Relational Databases](https://graphics.stanford.edu/papers/polaris_extended/polaris.pdf)

-   [Tableau Technology](https://www.tableau.com/products/technology)

-   [Show Me: Automatic Presentation for Visual Analysis](https://www.tableau.com/sites/default/files/whitepapers/081027-infovis-showme-vfinal-fix.pdf)

-   [Whitepaper: Visual Analysis for Everyone](https://www.tableau.com/sites/default/files/whitepapers/TS_WP_VA_F2_new_temp_0.pdf)

-   [Analytics Best Practices in Tableau](https://help.tableau.com/current/blueprint/en-us/bp_analytics_best_practices_in_tableau.htm)

**Learning Tableau**

-   [Getting Started](https://www.tableau.com/learn/tutorials/on-demand/getting-started)

-   [Tableau Interface](http://www.tableau.com/learn/tutorials/on-demand/tableau-interface)

-   [Getting Started with Visual Analytics](https://www.tableau.com/learn/tutorials/on-demand/getting-started-visual-analytics)

-   [Formatting](https://www.tableau.com/learn/tutorials/on-demand/formatting)

-   [The Formatting Pane](https://www.tableau.com/learn/tutorials/on-demand/formatting-pane)

-   [Reference lines](https://www.tableau.com/learn/tutorials/on-demand/reference-lines)

-   [Getting Started with Data](https://www.tableau.com/learn/tutorials/on-demand/getting-started-data) also self-learn the remaining videos under ’‘’Connecting to Data’’’

-   [Getting Started with Tableau Prep Builder](https://www.tableau.com/learn/tutorials/on-demand/getting-started-tableau-prep). This is a brand-new product from Tableau designed to help everyone quickly and confidently combine, shape, and clean their data for analysis. You can learn more about Tableau Data Prep [here](https://www.tableau.com/about/blog/2018/3/tableau-prep-project-maestro-here-redefining-your-data-prep-experience-84620). You are encouraged to view all the 12 videos under ’‘’Tableau Prep’’’.
