---
title: "Hands-on 9.1 & 9.2"
subtitle: " "
author: "Guan Jhen LIN"
date: "June 11 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

Hands on 9 consists of 5 different exercises, topics covered in this exercise:

-   [Building Ternary Plot with R](https://r4va.netlify.app/chap13).

-   [Visual Correlation Analysis](https://r4va.netlify.app/chap06).

# 9.1 Creating Ternary Plot with R

Ternary plots are a way of displaying the distribution and variability of three-part compositional data. (For example, the proportion of aged, economy active and young population or sand, silt, and clay in soil.) It’s display is a triangle with sides scaled from 0 to 1. Each side represents one of the three components. A point is plotted so that a line drawn perpendicular from the point to each leg of the triangle intersect at the component values of the point.

```{r}
pacman::p_load(plotly, ggtern, tidyverse)
```

## 9.1.1 Load Data

```{r}
pop_data <- read_csv("data/respopagsex2000to2018_tidy.csv") 
```

```{r}
#Deriving the young, economy active and old measures
agpop_mutated <- pop_data %>%
  mutate(`Year` = as.character(Year))%>%
  spread(AG, Population) %>%
  mutate(YOUNG = rowSums(.[4:8]))%>%
  mutate(ACTIVE = rowSums(.[9:16]))  %>%
  mutate(OLD = rowSums(.[17:21])) %>%
  mutate(TOTAL = rowSums(.[22:24])) %>%
  filter(Year == 2018)%>%
  filter(TOTAL > 0)
```

## 9.1.2 Plotting Ternary Diagram with R

```{r}
ggtern(data=agpop_mutated,aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point()
```

```{r}
ggtern(data=agpop_mutated, aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point() +
  labs(title="Population structure, 2015") +
  theme_rgbw()
```

## 9.1.3 Plotting an interative ternary diagram

The code below create an interactive ternary plot using plot_ly() function of Plotly R.

-   reusable function for creating annotation object

```{r}
label <- function(txt) {
  list(
    text = txt, 
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper", 
    align = "center",
    font = list(family = "serif", size = 15, color = "white"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}

# reusable function for axis formatting
axis <- function(txt) {
  list(
    title = txt, tickformat = ".0%", tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"), 
  baxis = axis("Active"), 
  caxis = axis("Old")
)

# Initiating a plotly visualization 
plot_ly(
  agpop_mutated, 
  a = ~YOUNG, 
  b = ~ACTIVE, 
  c = ~OLD, 
  color = I("black"), 
  type = "scatterternary"
) %>%
  layout(
    annotations = label("Ternary Markers"), 
    ternary = ternaryAxes
  )
```

# 9.2 Visual Correlation Analysis

Correlation coefficient is a popular statistic that use to measure the type and strength of the relationship between two variables. The values of a correlation coefficient ranges between -1.0 and 1.0. A correlation coefficient of 1 shows a perfect linear relationship between the two variables, while a -1.0 shows a perfect inverse relationship between the two variables. A correlation coefficient of 0.0 shows no linear relationship between the two variables.

When multivariate data are used, the correlation coefficeints of the pair comparisons are displayed in a table form known as correlation matrix or scatterplot matrix.

There are three broad reasons for computing a correlation matrix.

-   To reveal the relationship between high-dimensional variables pair-wisely.

-   To input into other analyses. For example, people commonly use correlation matrices as inputs for exploratory factor analysis, confirmatory factor analysis, structural equation models, and linear regression when excluding missing values pairwise.

-   As a diagnostic when checking other analyses. For example, with linear regression a high amount of correlations suggests that the linear regression’s estimates will be unreliable.

When the data is large, both in terms of the number of observations and the number of variables, [Corrgram](http://www.datavis.ca/papers/corrgram.pdf) tend to be used to visually explore and analyse the structure and the patterns of relations among variables. It is designed based on two main schemes:

-   Rendering the value of a correlation to depict its sign and magnitude, and

-   Reordering the variables in a correlation matrix so that “similar” variables are positioned adjacently, facilitating perception.

```{r}
pacman::p_load(corrplot, ggstatsplot, tidyverse)
```

```{r}
wine <- read_csv("data/wine_quality.csv")
```

## 9.2.1 Building Correlation Matrix: pairs() method

There are more than one way to build scatterplot matrix with R. In this section, you will learn how to create a scatterplot matrix by using the pairs function of R Graphics.

Before you continue to the next step, you should read the syntax description of pairsfunction.

Building a basic correlation matrix Figure below shows the scatter plot matrix of Wine Quality Data. It is a 11 by 11 matrix.

```{r}
pairs(wine[,1:11])
```

The required input of pairs() can be a matrix or data frame. The code chunk used to create the scatterplot matrix is relatively simple. It uses the default pairs function. Columns 2 to 12 of wine dataframe is used to build the scatterplot matrix. The variables are: fixed acidity, volatile acidity, citric acid, residual sugar, chlorides, free sulfur dioxide, total sulfur dioxide, density, pH, sulphates and alcohol.

```{r}
pairs(wine[,2:12])
```

## 9.2.2 Drawing the lower corner

pairs function of R Graphics provided many customisation arguments. For example, it is a common practice to show either the upper half or lower half of the correlation matrix instead of both. This is because a correlation matrix is symmetric.

To show the lower half of the correlation matrix, the upper.panel argument will be used as shown in the code chunk below.

```{r}
pairs(wine[,2:12], upper.panel = NULL)
```

Similarly, you can display the upper half of the correlation matrix by using the code chun below.

```{r}
pairs(wine[,2:12], lower.panel = NULL)
```

## 9.2.3 Including with correlation coefficients

To show the correlation coefficient of each pair of variables instead of a scatter plot, panel.cor function will be used. This will also show higher correlations in a larger font.

Don’t worry about the details for now-just type this code into your R session or script. Let’s have more fun way to display the correlation matrix.

```{r}
panel.cor <- function(x, y, digits=2, prefix="", cex.cor, ...) {
usr <- par("usr")
on.exit(par(usr))
par(usr = c(0, 1, 0, 1))
r <- abs(cor(x, y, use="complete.obs"))
txt <- format(c(r, 0.123456789), digits=digits)[1]
txt <- paste(prefix, txt, sep="")
if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
text(0.5, 0.5, txt, cex = cex.cor * (1 + r) / 2)
}

pairs(wine[,2:12], 
      upper.panel = panel.cor)
```

## 9.2.4 Visualising Correlation Matrix: ggcormat()

One of the major limitation of the correlation matrix is that the scatter plots appear very cluttered when the number of observations is relatively large (i.e. more than 500 observations). To over come this problem, **Corrgram** data visualisation technique suggested by D. J. Murdoch and E. D. Chow (1996) and Friendly, M (2002) and will be used.

The are at least three R packages provide function to plot corrgram, they are:

-   [corrgram](https://cran.r-project.org/web/packages/corrgram/)

-   [ellipse](https://cran.r-project.org/web/packages/ellipse/)

-   [corrplot](https://cran.r-project.org/web/packages/corrplot/)

On top that, some R package like ggstatsplot package also provides functions for building corrgram.

In this section, you will learn how to visualising correlation matrix by using [*ggcorrmat()*](https://indrajeetpatil.github.io/ggstatsplot/reference/ggcorrmat.html) of [**ggstatsplot**](https://indrajeetpatil.github.io/ggstatsplot/) package.

## 9.2.5 The basic plot

On of the advantage of using ggcorrmat() over many other methods to visualise a correlation matrix is it’s ability to provide a comprehensive and yet professional statistical report as shown in the figure below.

```{r}
ggstatsplot::ggcorrmat(
  data = wine, 
  cor.vars = 1:11)
```

```{r}
ggstatsplot::ggcorrmat(
  data = wine, 
  cor.vars = 1:11,
  ggcorrplot.args = list(outline.color = "black", 
                         hc.order = TRUE,
                         tl.cex = 10),
  title    = "Correlogram for wine dataset",
  subtitle = "Four pairs are no significant at p < 0.05"
)
```

Things to learn from the code chunk above:

-   `cor.vars` argument is used to compute the correlation matrix needed to build the corrgram.

-   `ggcorrplot.args` argument provide additional (mostly aesthetic) arguments that will be passed to [`ggcorrplot::ggcorrplot`](http://www.sthda.com/english/wiki/ggcorrplot-visualization-of-a-correlation-matrix-using-ggplot2) function. The list should avoid any of the following arguments since they are already internally being used: `corr`, `method`, `p.mat`, `sig.level`, `ggtheme`, `colors`, `lab`, `pch`, `legend.title`, `digits`.

The sample sub-code chunk can be used to control specific component of the plot such as the font size of the x-axis, y-axis, and the statistical report.

```{r}
ggplot.component = list(
    theme(text=element_text(size=5),
      axis.text.x = element_text(size = 8),
      axis.text.y = element_text(size = 8)))
```

## 9.2.6 Building multiple plots

Since ggstasplot is an extension of ggplot2, it also supports faceting. However the feature is not available in *ggcorrmat()* but in the [*grouped_ggcorrmat()*](https://indrajeetpatil.github.io/ggstatsplot/reference/grouped_ggcorrmat.html) of **ggstatsplot**.

```{r}

grouped_ggcorrmat(
  data = wine,
  cor.vars = 1:11,
  grouping.var = type,
  type = "robust",
  p.adjust.method = "holm",
  plotgrid.args = list(ncol = 2),
  ggcorrplot.args = list(outline.color = "black", 
                         hc.order = TRUE,
                         tl.cex = 10),
  annotation.args = list(
    tag_levels = "a",
    title = "Correlogram for wine dataset",
    subtitle = "The measures are: alcohol, sulphates, fixed acidity, citric acid, chlorides, residual sugar, density, free sulfur dioxide and volatile acidity",
    caption = "Dataset: UCI Machine Learning Repository"
  )
)

```

Things to learn from the code chunk above:

-   to build a facet plot, the only argument needed is `grouping.var`.

-   Behind *group_ggcorrmat()*, **patchwork** package is used to create the multiplot. `plotgrid.args` argument provides a list of additional arguments passed to [*patchwork::wrap_plots*](https://patchwork.data-imaginist.com/reference/wrap_plots.html), except for guides argument which is already separately specified earlier.

-   Likewise, `annotation.args` argument is calling [*plot annotation arguments*](https://patchwork.data-imaginist.com/reference/plot_annotation.html) of patchwork package.

## 9.2.7 Visualising Correlation Matrix using corrplot Package

```{r}
wine.cor <- cor(wine[, 1:11])
```

```{r}
corrplot(wine.cor)
```

Notice that the default visual object used to plot the corrgram is circle. The default layout of the corrgram is a symmetric matrix. The default colour scheme is diverging blue-red. Blue colours are used to represent pair variables with positive correlation coefficients and red colours are used to represent pair variables with negative correlation coefficients. The intensity of the colour or also know as saturation is used to represent the strength of the correlation coefficient. Darker colours indicate relatively stronger linear relationship between the paired variables. On the other hand, lighter colours indicates relatively weaker linear relationship.

## 9.2.8 Working with visual geometrics

In corrplot package, there are seven visual geometrics (parameter method) can be used to encode the attribute values. They are: circle, square, ellipse, number, shade, color and pie. The default is circle. As shown in the previous section, the default visual geometric of corrplot matrix is circle. However, this default setting can be changed by using the method argument as shown in the code chunk below.

```{r}
corrplot(wine.cor, 
         method = "ellipse") 
```

## 9.2.9 Working with layout

corrplor() supports three layout types, namely: “full”, “upper” or “lower”. The default is “full” which display full matrix. The default setting can be changed by using the type argument of corrplot().

```{r}
corrplot(wine.cor, 
         method = "ellipse", 
         type="lower")
```

The default layout of the corrgram can be further customised. For example, arguments diag and tl.col are used to turn off the diagonal cells and to change the axis text label colour to black colour respectively as shown in the code chunk and figure below.

```{r}
corrplot(wine.cor, 
         method = "ellipse", 
         type="lower",
         diag = FALSE,
         tl.col = "black")

```

## 9.2.10 Working with mixed layout

With corrplot package, it is possible to design corrgram with mixed visual matrix of one half and numerical matrix on the other half. In order to create a coorgram with mixed layout, the corrplot.mixed().
