---
title: "Hands-on Exercise 4"
subtitle: ""
author: "Guan Jhen Lin"
date: "May 1, 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
output:
  pdf_document:
    latex_engine: xelatex
---

# 4.1 Visual Statistical Analysis

Getting Started

```{r}
pacman::p_load(ggstatsplot, tidyverse)
```

```{r}
exam <- read_csv("data/Exam_data.csv")
```

One-sample test: gghistostats() method

```{r}
set.seed(1234)

gghistostats(
  data = exam,
  x = ENGLISH,
  type = "bayes",
  test.value = 60,
  xlab = "English scores"
)
```

-   A Bayes factor is the ratio of the likelihood of one particular hypothesis to the likelihood of another. It can be interpreted as a measure of the strength of evidence in favor of one theory among two competing theories.

-   That’s because the Bayes factor gives us a way to evaluate the data in favor of a null hypothesis, and to use external information to do so. It tells us what the weight of the evidence is in favor of a given hypothesis.

Learn how to interpret Bayes Factor [Here](https://www-tandfonline-com.libproxy.smu.edu.sg/doi/pdf/10.1080/00031305.1999.10474443?needAccess=true).

Two-sample mean test: ggbetweenstats()

```{r}

ggbetweenstats(
  data = exam,
  x = GENDER, 
  y = MATHS,
  type = "np",
  messages = FALSE
)
```

Oneway ANOVA Test: ggbetweenstats() method

```{r}
ggbetweenstats(
  data = exam,
  x = RACE, 
  y = ENGLISH,
  type = "p",
  mean.ci = TRUE, 
  pairwise.comparisons = TRUE, 
  pairwise.display = "s",
  p.adjust.method = "fdr",
  messages = FALSE
)
```

Significant Test of Correlation: ggscatterstats()

```{r}
ggscatterstats(
  data = exam,
  x = MATHS,
  y = ENGLISH,
  marginal = FALSE,
  )
```

Significant Test of Association (Depedence) : ggbarstats() methods

```{r}
exam1 <- exam %>% 
  mutate(MATHS_bins = 
           cut(MATHS, 
               breaks = c(0,60,75,85,100))
)
```

using ggbarstats

```{r}
ggbarstats(exam1, 
           x = MATHS_bins, 
           y = GENDER)

```

# 4.1.1 Visualising Models

```{r}
pacman::p_load(readxl, performance, parameters, see)
```

```{r}
car_resale <- read_xls("data/ToyotaCorolla.xls", 
                       "data")
car_resale
```

Multiple Regression Model using lm()

```{r}
model <- lm(Price ~ Age_08_04 + Mfg_Year + KM + 
              Weight + Guarantee_Period, data = car_resale)
model
```

Model Diagnostic: checking for multicolinearity

```{r}
check_collinearity(model)
```

```{r}
check_c <- check_collinearity(model)
plot(check_c)
```

Model Diagnostic: checking normality assumption

```{r}
model1 <- lm(Price ~ Age_08_04 + KM + 
              Weight + Guarantee_Period, data = car_resale)
```

```{r}
check_n <- check_normality(model1)
```

```{r}
plot(check_n)
```

Model Diagnostic: Check model for homogeneity of variances

```{r}
check_h <- check_heteroscedasticity(model1)
```

```{r}
plot(check_h)
```

Model Diagnostic: Complete check

```{r}
check_model(model1)
```

Visualising Regression Parameters: see methods

```{r}
plot(parameters(model1))
```

Visualising Regression Parameters: ggcoefstats() methods

```{r}
ggcoefstats(model1, 
            output = "plot")
```

# 4.2 Visualising Uncertainty

```{r}
devtools::install_github("wilkelab/ungeviz")
```

```{r}
pacman::p_load(ungeviz, plotly, crosstalk,
               DT, ggdist, ggridges,
               colorspace, gganimate, tidyverse)
```

```{r}
exam <- read_csv("data/Exam_data.csv")
```

Visualizing the uncertainty of point estimates: ggplot2 methods

Important: Don’t confuse the uncertainty of a point estimate with the variation in the sample

```{r}
my_sum <- exam %>%
  group_by(RACE) %>%
  summarise(
    n=n(),
    mean=mean(MATHS),
    sd=sd(MATHS)
    ) %>%
  mutate(se=sd/sqrt(n-1))
```

-   group_by() of dplyr package is used to group the observation by RACE,
-   summarise() is used to compute the count of observations, mean, standard deviation
-   mutate() is used to derive standard error of Maths by RACE, and
-   the output is save as a tibble data table called my_sum.

```{r}
knitr::kable(head(my_sum), format = 'html')
```

Plotting standard error bars of point estimates

```{r}
ggplot(my_sum) +
  geom_errorbar(
    aes(x=RACE, 
        ymin=mean-se, 
        ymax=mean+se), 
    width=0.2, 
    colour="black", 
    alpha=0.9, 
    size=0.5) +
  geom_point(aes
           (x=RACE, 
            y=mean), 
           stat="identity", 
           color="red",
           size = 1.5,
           alpha=1) +
  ggtitle("Standard error of mean maths score by rac")
```

Plotting confidence interval of point estimates

```{r}
ggplot(my_sum) +
  geom_errorbar(
    aes(x=reorder(RACE, -mean), 
        ymin=mean-1.96*se, 
        ymax=mean+1.96*se), 
    width=0.2, 
    colour="black", 
    alpha=0.9, 
    size=0.5) +
  geom_point(aes
           (x=RACE, 
            y=mean), 
           stat="identity", 
           color="red",
           size = 1.5,
           alpha=1) +
  labs(x = "Maths score",
       title = "95% confidence interval of mean maths score by race")
```

Visualizing the uncertainty of point estimates with interactive error bars

```{r}

shared_df = SharedData$new(my_sum)

bscols(widths = c(4,8),
       ggplotly((ggplot(shared_df) +
                   geom_errorbar(aes(
                     x=reorder(RACE, -mean),
                     ymin=mean-2.58*se, 
                     ymax=mean+2.58*se), 
                     width=0.2, 
                     colour="black", 
                     alpha=0.9, 
                     size=0.5) +
                   geom_point(aes(
                     x=RACE, 
                     y=mean, 
                     text = paste("Race:", `RACE`, 
                                  "<br>N:", `n`,
                                  "<br>Avg. Scores:", round(mean, digits = 2),
                                  "<br>95% CI:[", 
                                  round((mean-2.58*se), digits = 2), ",",
                                  round((mean+2.58*se), digits = 2),"]")),
                     stat="identity", 
                     color="red", 
                     size = 1.5, 
                     alpha=1) + 
                   xlab("Race") + 
                   ylab("Average Scores") + 
                   theme_minimal() + 
                   theme(axis.text.x = element_text(
                     angle = 45, vjust = 0.5, hjust=1)) +
                   ggtitle("99% Confidence interval of average /<br>maths scores by race")), 
                tooltip = "text"), 
       DT::datatable(shared_df, 
                     rownames = FALSE, 
                     class="compact", 
                     width="100%", 
                     options = list(pageLength = 10,
                                    scrollX=T), 
                     colnames = c("No. of pupils", 
                                  "Avg Scores",
                                  "Std Dev",
                                  "Std Error")) %>%
         formatRound(columns=c('mean', 'sd', 'se'),
                     digits=2))

```

Visualising Uncertainty: ggdist package

-   [**ggdist**](https://mjskay.github.io/ggdist/) is an R package that provides a flexible set of ggplot2 geoms and stats designed especially for visualising distributions and uncertainty.

-   It is designed for both frequentist and Bayesian uncertainty visualization, taking the view that uncertainty visualization can be unified through the perspective of distribution visualization:

    -   for frequentist models, one visualises confidence distributions or bootstrap distributions (see vignette(“freq-uncertainty-vis”));

    -   for Bayesian models, one visualises probability distributions (see the tidybayes package, which builds on top of ggdist).

Visualizing the uncertainty of point estimates: ggdist methods

In the code chunk below, stat_pointinterval() of ggdist is used to build a visual for displaying distribution of maths scores by race.

```{r}
exam %>%
  ggplot(aes(x = RACE, 
             y = MATHS)) +
  stat_pointinterval() +
  labs(
    title = "Visualising confidence intervals of mean math score",
    subtitle = "Mean Point + Multiple-interval plot")
```

For example, in the code chunk below the following arguments are used:

.width = 0.95 .point = median .interval = qi

```{r}
exam %>%
  ggplot(aes(x = RACE, y = MATHS)) +
  stat_pointinterval(.width = 0.95,
  .point = median,
  .interval = qi) +
  labs(
    title = "Visualising confidence intervals of median math score",
    subtitle = "Median Point + Multiple-interval plot")
```

Visualizing the uncertainty of point estimates: ggdist methods

```{r}
exam %>%
  ggplot(aes(x = RACE, 
             y = MATHS)) +
  stat_pointinterval(
    show.legend = FALSE) +   
  labs(
    title = "Visualising confidence intervals of mean math score",
    subtitle = "Mean Point + Multiple-interval plot")
```

Visualizing the uncertainty of point estimates: ggdist methods

In the code chunk below, stat_gradientinterval() of ggdist is used to build a visual for displaying distribution of maths scores by race.

```{r}
exam %>%
  ggplot(aes(x = RACE, 
             y = MATHS)) +
  stat_gradientinterval(   
    fill = "skyblue",      
    show.legend = TRUE     
  ) +                        
  labs(
    title = "Visualising confidence intervals of mean math score",
    subtitle = "Gradient + interval plot")
```

Visualising Uncertainty with Hypothetical Outcome Plots (HOPs)

```{r}
devtools::install_github("wilkelab/ungeviz")
```

```{r}
library(ungeviz)
```

```{r}
ggplot(data = exam, 
       (aes(x = factor(RACE), y = MATHS))) +
  geom_point(position = position_jitter(
    height = 0.3, width = 0.05), 
    size = 0.4, color = "#0072B2", alpha = 1/2) +
  geom_hpline(data = sampler(25, group = RACE), height = 0.6, color = "#D55E00") +
  theme_bw() + 
  # `.draw` is a generated column indicating the sample draw
  transition_states(.draw, 1, 3)
```

Visualising Uncertainty with Hypothetical Outcome Plots (HOPs)

```{r}
ggplot(data = exam, 
       (aes(x = factor(RACE), 
            y = MATHS))) +
  geom_point(position = position_jitter(
    height = 0.3, 
    width = 0.05), 
    size = 0.4, 
    color = "#0072B2", 
    alpha = 1/2) +
  geom_hpline(data = sampler(25, 
                             group = RACE), 
              height = 0.6, 
              color = "#D55E00") +
  theme_bw() + 
  transition_states(.draw, 1, 3)
```

# 4.3 Funnel Plots for Fair Comparisons

Funnel plot is a specially designed data visualisation for conducting unbiased comparison between outlets, stores or business entities. By the end of this hands-on exercise, you will gain hands-on experience on:

-   plotting funnel plots by using **funnelPlotR** package,

-   plotting static funnel plot by using ggplot2 package, and

-   plotting interactive funnel plot by using both **plotly R** and **ggplot2** packages.

```{r}
pacman::p_load(tidyverse, FunnelPlotR, plotly, knitr)
```

```{r}
library(FunnelPlotR)  

```

```{r}
covid19 <- read_csv("data/COVID-19_DKI_Jakarta.csv") %>%

  mutate_if(is.character, as.factor)
```

```{r}
funnel_plot(
  .data = covid19,
  numerator = Positive,
  denominator = Death,
  group = `Sub-district`
)
```

```         
A funnel plot object with 267 points of which 0 are outliers.  Plot is adjusted for overdispersion. 
```

Things to learn from the code chunk above.

-   `group` in this function is different from the scatter plot. Here, it defines the level of the points to be plotted i.e. Sub-district, District or City. If Cityc is chosen, there are only six data points.

-   By default, `data_type`argument is “SR”.

-   `limit`: Plot limits, accepted values are: 95 or 99, corresponding to 95% or 99.8% quantiles of the distribution.

FunnelPlotR methods: Makeover 1

```{r}
funnel_plot(
  .data = covid19,
  numerator = Death,
  denominator = Positive,
  group = `Sub-district`,
  data_type = "PR",     #<<
  xrange = c(0, 6500),  #<<
  yrange = c(0, 0.05)   #<<
)
```

```         
A funnel plot object with 267 points of which 7 are outliers.  Plot is adjusted for overdispersion. 
```

Things to learn from the code chunk above. + `data_type` argument is used to change from default “SR” to “PR” (i.e.proportions). + `xrange` and `yrange` are used to set the range of x-axis and y-axis

FunnelPlotR methods: Makeover 2

```{r}
funnel_plot(
  .data = covid19,
  numerator = Death,
  denominator = Positive,
  group = `Sub-district`,
  data_type = "PR",   
  xrange = c(0, 6500),  
  yrange = c(0, 0.05),
  label = NA,
  title = "Cumulative COVID-19 Fatality Rate by Cumulative Total Number of COVID-19 Positive Cases", #<<           
  x_label = "Cumulative COVID-19 Positive Cases", #<<
  y_label = "Cumulative Fatality Rate"  #<<
)
```

```        
A funnel plot object with 267 points of which 7 are outliers. Plot is adjusted for overdispersion. 
``` 

Things to learn from the code chunk above.

-   `label = NA` argument is to removed the default label outliers feature.

-   `title` argument is used to add plot title.

-   `x_label` and `y_label` arguments are used to add/edit x-axis and y-axis titles.

Funnel Plot for Fair Visual Comparison: ggplot2 methods

Computing the basic derived fields

```{r}
funnel_plot(
  .data = covid19, 
  numerator = "Positive", 
  denominator = "Death", 
  group = "Sub-district"
  )

```

```{r}
df <- covid19 %>%
  mutate(rate = Death / Positive) %>%
  mutate(rate.se = sqrt((rate*(1-rate)) / (Positive))) %>%
  filter(rate > 0)
```

```{r}
fit.mean <- weighted.mean(df$rate, 1/df$rate.se^2)
```

Calculate lower and upper limits for 95% and 99.9% CI

```{r}
number.seq <- seq(1, max(df$Positive), 1)
number.ll95 <- fit.mean - 1.96 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
number.ul95 <- fit.mean + 1.96 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
number.ll999 <- fit.mean - 3.29 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
number.ul999 <- fit.mean + 3.29 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
dfCI <- data.frame(number.ll95, number.ul95, number.ll999, 
                   number.ul999, number.seq, fit.mean)
```

Plotting a static funnel plot

```{r}
p <- ggplot(df, aes(x = Positive, y = rate)) +
  geom_point(aes(label=`Sub-district`), 
             alpha=0.4) +
  geom_line(data = dfCI, 
            aes(x = number.seq, 
                y = number.ll95), 
            size = 0.4, 
            colour = "grey40", 
            linetype = "dashed") +
  geom_line(data = dfCI, 
            aes(x = number.seq, 
                y = number.ul95), 
            size = 0.4, 
            colour = "grey40", 
            linetype = "dashed") +
  geom_line(data = dfCI, 
            aes(x = number.seq, 
                y = number.ll999), 
            size = 0.4, 
            colour = "grey40") +
  geom_line(data = dfCI, 
            aes(x = number.seq, 
                y = number.ul999), 
            size = 0.4, 
            colour = "grey40") +
  geom_hline(data = dfCI, 
             aes(yintercept = fit.mean), 
             size = 0.4, 
             colour = "grey40") +
  coord_cartesian(ylim=c(0,0.05)) +
  annotate("text", x = 1, y = -0.13, label = "95%", size = 3, colour = "grey40") + 
  annotate("text", x = 4.5, y = -0.18, label = "99%", size = 3, colour = "grey40") + 
  ggtitle("Cumulative Fatality Rate by Cumulative Number of COVID-19 Cases") +
  xlab("Cumulative Number of COVID-19 Cases") + 
  ylab("Cumulative Fatality Rate") +
  theme_light() +
  theme(plot.title = element_text(size=12),
        legend.position = c(0.91,0.85), 
        legend.title = element_text(size=7),
        legend.text = element_text(size=7),
        legend.background = element_rect(colour = "grey60", linetype = "dotted"),
        legend.key.height = unit(0.3, "cm"))
p
```

Interactive Funnel Plot: plotly + ggplot2

```{r}
fp_ggplotly <- ggplotly(p,
  tooltip = c("label", 
              "x", 
              "y"))
fp_ggplotly


```

## **12.6 References**

-   [funnelPlotR](https://nhs-r-community.github.io/FunnelPlotR/) package.

-   [Funnel Plots for Indirectly-standardised ratios](https://nhs-r-community.github.io/FunnelPlotR/articles/funnel_plots.html).

-   [Changing funnel plot options](https://nhs-r-community.github.io/FunnelPlotR/articles/changing_funnel_plot_options.html)

-   [ggplot2](https://ggplot2.tidyverse.org/) package.
